---
title: "Data Analysis and Methods"
authors: "Jacob Posner, Evan Burns, Bowman Wingard, Cameron Latvis"
format: 
  html:
    embed-resources: true
    toc: true
    theme: cerulean
---

# Data

We collected data from Retrosheet, a database that tracks and stores Major League Baseball data at the game level. We gathered data on each regular season baseball game from 1995 to 2023. Each game’s data contains variables such as the date, team information, stadium information, attendance, score, team stats, umpire information, and pitcher information. This is important because the data gives us an abundance of information to match or weigh on if we choose to go down those paths. The dimensions of our dataset are 68185 observations of 175 variables. 

We performed several filters from here to produce the final data set we will use for our analysis. First, we only included games were included in the regular season. We made a new data set for each individual team that we then can analyze and compare to each other. We made some adjustments by team as well. For example, we are analyzing the years 1995 and after, however, the Tampa Bay Rays became a team in 1998 so we only could gather data starting that year. Another adjustment we made is to treat the Florida Marlins and the Miami Marlins as the same team. They are the same franchise but were renamed in 2012. In the data, the two names are coded differently and are working around it by naming both the Miami Marlins. We also made some other changes to help make it easier to add confounding variables. We added binary categories for opposing team statistics that can help us control for certain variables. 

Lastly, we chose to exclude the years 2020 and 2021 from our analysis. Since we are concerned with home-field advantage, we anticipate a portion of home-field advantage to stem from fans attending games and making their presence felt. Because of the pandemic, there was extremely limited attendance in these two years, which is why they are excluded in our data sets. We have a total of 30 data sets; one for each MLB team, and will analyze them individually and compare them to each other. 

For the purposes of this milestone, we are conducting an analysis on 1 team. We hope to write a function that will complete the analysis for us so we don't have to copy and paste code 30 different times. 

# Dag Plot 

```{r}
library(dagitty)

# Define the DAG
dag <- dagitty('
dag {
  bb="-8,-6,10,0"
  "Day/Night" [pos="2.887,-4.914"]
  "Home (treatment)" [exposure,pos="-6.498,-4.544"]
  "Opponent Skill" [pos="5.682,-4.867"]
  "Opponent Starter" [pos="7.232,-4.833"]
  Attendance [pos="0.745,-4.633"]
  Comfort [pos="-2.499,-4.360"]
  DOW [pos="4.315,-4.915"]
  Umpire [pos="8.232,-4.873"]
  Win [outcome,pos="7.191,-4.503"]
  performance [pos="1.153,-4.174"]

  "Day/Night" -> Attendance
  "Home (treatment)" -> Comfort
  "Opponent Skill" -> Win
  "Opponent Starter" -> Win
  Attendance -> Comfort
  Comfort -> Win
  Comfort -> performance
  DOW -> Attendance
  Umpire -> Win
  performance -> Win
}
')

# Plot the DAG
plot(dag)

```


```{r include = FALSE}
knitr::opts_chunk$set(
  warning = FALSE,
  message = FALSE,
  collapse = TRUE, 
  echo = TRUE, 
  fig.height = 5, 
  fig.width = 7,
  fig.align = 'center')
```

```{r}
library(dplyr)
library(tidyverse)
library(ggplot2)
library(tidyverse)
library(broom)
library(MatchIt)
library(marginaleffects)
library(dagitty)
library(scales)
```

```{r}
#Load the Data
ANA<-read_csv("Data/Team Datasets/ANA_games.csv")

ANA %>% 
  dim()
ANA %>% 
  names()
```

## ATT

```{r}
match_out_full_ANA <- matchit(
    Home ~ Opp_Win_P_Cat + Opp_Playoffs + Opp_WS + Dow + Day_Night,
    data = ANA,
    method = "full",
    exact = ~ Opp_Win_P_Cat + Opp_Playoffs + Opp_WS + Dow + Day_Night,  # Example of exact matching variables
    distance = "glm",
    estimand = "ATT"
)
```

```{r}
match_out_full_summ_ANA <- summary(match_out_full_ANA, interactions = TRUE)
```

```{r, fig.height= 15}
plot(match_out_full_summ_ANA)
```

```{r}
match_data_full_ATT_ANA <- match.data(match_out_full_ANA)

mod2 <- lm(ANA_win ~ Home * (Opp_Win_P_Cat + Opp_Playoffs + Opp_WS + Attendance + Dow + Day_Night), data = match_data_full_ATT_ANA, weights = weights)

avg_comparisons(
    mod2,
    variables = "Home",
    vcov = ~subclass,
    newdata = filter(match_data_full_ATT_ANA, Home == 1))
```

For all Angle's home games, if they had been played away, we would expect their winning percentage to be 9.4% less. This equates to about 7.6 games, meaning the Angels would've lost about 8 games more if they had played every home game as an away game. 

## ATC

```{r}
match_out_full_ATC_ANA <- matchit(
    Home ~ Opp_Win_P_Cat + Opp_Playoffs + Opp_WS + Dow + Day_Night,
    data = ANA,
    method = "full",
    exact = ~ Opp_Win_P_Cat + Opp_Playoffs + Opp_WS + Dow + Day_Night,
    distance = "glm",
    estimand = "ATC"
)
```

```{r}
match_out_full_summ_ATC_ANA <- summary(match_out_full_ATC_ANA, interactions = TRUE)
match_out_full_summ_ATC_ANA
```

```{r, fig.height= 15}
plot(match_out_full_summ_ATC_ANA)
```

```{r}
match_data_full_ATC_ANA <- match.data(match_out_full_ATC_ANA)

mod3 <- lm(ANA_win ~ Home * (Opp_Win_P_Cat + Opp_Playoffs + Opp_WS + Dow + Day_Night), data = match_data_full_ATC_ANA, weights = weights)

avg_comparisons(
    mod3,
    variables = "Home",
    vcov = ~subclass,
    newdata = filter(match_data_full_ATC_ANA, Home == 0))
```

For all Angel's away games, if they had been played at home, we would expect their winning percentage to be 6.6% greater. This equates to about 5.3 games, meaning the Angels would've won about 5 games more if they had played every away game as a home game. 

```{r}
match_data_full_ANA <- match.data(match_out_full_ANA)

mod4 <- lm(ANA_win ~ Home * (Opp_Win_P_Cat + Opp_Playoffs + Opp_WS + Dow + Day_Night), data = match_data_full_ANA, weights = weights)

avg_comparisons(
    mod4,
    variables = "Home",
    vcov = ~subclass)
```

> If the Angels played every game away, we would expect the Angels win percentage to be 8.6% lower than if they played every game at home. This equates to about 13.9 games which means the Angels would win 14 games more if they played every game at home versus every game away. This is very interesting considering this could be the difference between winning your division and not making the playoffs. 

