---
title: "SensAnalysis2"
format: html
---

# Load Packages
```{r}
library(readr)
library(dplyr)
library(EValue)
```

# Import Datasets / Make a List
```{r}
# Create a list of team datasets that can be iterated through
team_files <- list(
  ANA = "Data/Team Datasets/ANA_games.csv",
  DET = "Data/Team Datasets/DET_games.csv",
  KCA = "Data/Team Datasets/KCA_games.csv",
  PHI = "Data/Team Datasets/PHI_games.csv",
  WAS = "Data/Team Datasets/WAS_games.csv",
  BOS = "Data/Team Datasets/BOS_games.csv",
  CIN = "Data/Team Datasets/CIN_games.csv",
  SDN = "Data/Team Datasets/SDN_games.csv",
  SFN = "Data/Team Datasets/SFN_games.csv",
  SLN = "Data/Team Datasets/SLN_games.csv",
  CLE = "Data/Team Datasets/CLE_games.csv",
  HOU = "Data/Team Datasets/HOU_games.csv",
  MIA = "Data/Team Datasets/MIA_games.csv",
  MIL = "Data/Team Datasets/MIL_games.csv",
  ARI = "Data/Team Datasets/ARI_games.csv",
  NYA = "Data/Team Datasets/NYA_games.csv",
  OAK = "Data/Team Datasets/OAK_games.csv",
  SEA = "Data/Team Datasets/SEA_games.csv",
  TBA = "Data/Team Datasets/TBA_games.csv",
  TEX = "Data/Team Datasets/TEX_games.csv",
  ATL = "Data/Team Datasets/ATL_games.csv",
  BAL = "Data/Team Datasets/BAL_games.csv",
  CHA = "Data/Team Datasets/CHA_games.csv",
  MIN = "Data/Team Datasets/MIN_games.csv",
  TOR = "Data/Team Datasets/TOR_games.csv",
  CHN = "Data/Team Datasets/CHN_games.csv",
  COL = "Data/Team Datasets/COL_games.csv",
  LAN = "Data/Team Datasets/LAN_games.csv",
  NYN = "Data/Team Datasets/NYN_games.csv",
  PIT = "Data/Team Datasets/PIT_games.csv")
```

# Loop
```{r}
# Intialize a list to store the results
evalues_results <- list()

# Iterate over team abbreviations and their data files
for (abbr in names(team_files)) {
  # Read the data for the current team
  team_data <- read.csv(team_files[[abbr]])

  # Calculate n11, n10, n01, n00 for the current team
  n11 <- sum(team_data$Home == 1 & team_data[[paste0(abbr, "_win")]] == 1) # Home Game, Home Win
  n10 <- sum(team_data$Home == 1 & team_data[[paste0(abbr, "_win")]] == 0) # Home Game, Home Loss
  n01 <- sum(team_data$Home == 0 & team_data[[paste0(abbr, "_win")]] == 1) # Away Game, Away Win
  n00 <- sum(team_data$Home == 0 & team_data[[paste0(abbr, "_win")]] == 0) # Away Game, Away Loss
  
  # Compute E-value
  evalue <- evalues.RD(n11, n10, n01, n00, true = 0)
  
  # Store the result in the list
  evalues_results[[abbr]] <- list(
    home_win = n11,
    home_loss = n10,
    away_win = n01,
    away_loss = n00,
    evalue = evalue
  )
}
```

# Combining Results from the List

```{r}
# Combine results into a data frame
evalues_table <- do.call(rbind, lapply(names(evalues_results), function(abbr) {
  data.frame(
    Team = abbr,
    home_win = evalues_results[[abbr]]$home_win,
    home_loss = evalues_results[[abbr]]$home_loss,
    away_win = evalues_results[[abbr]]$away_win,
    away_loss = evalues_results[[abbr]]$away_loss,
    est_EValue = evalues_results[[abbr]]$evalue[1], # Extract the main E-value
    low_EValue = evalues_results[[abbr]]$evalue[2] # Extract the low E-value
  )
}))
```

# Results Arranged by E-Value - Higher is Better

```{r}
# Print the results
evalues_table %>% 
  arrange(desc(est.Evalue))

evalues_table %>% 
  summarise(mean(est.Evalue))
```

> Colorado has the highest e-value at around 2.1 which means that the degree to which our analysis could be affected by a cofounding variable is lower for them. This makes sense as well because the park is known to provide an advantage to hitters due to elevation and ball flight. Boston on the other hand has the lowest evalue at 1.44 which may suggest that our estimation and analysis for them is more sensitive to confounding variables. For boston specifically, their park layout with the green monster may confound in addition to other variables. The mean e-value for every team in the dataset is 1.6 which is not great and may suggest that there are uncounted for variables that influence the relationship between winning and playing at home. 
